Microbial interactions as drivers of a nitrification process in a chemostat

### 2020-05-25 tex/Article_Interactions.tex
finished POC and application.

Notations used throughout the article:
\begin{enumerate}
	\item $n$: the number of OTU considered.
	\item $n_i, \, i \in \{1,2\}$: the number of OTU in functional group $G_i$. In the example $G_1$ corresponds to Ammonium oxidizing bacteria (AOB) and $G_2$ corresponds to Ammonium oxidizing bacteria (NOB).
	\item Let $m$ be an interger then $[m]:=\{1,\dots,n \}$. 
	\item $x_i$ :  is the concentration of OTU $i$ measured in $[g/l]$. $i \in [n]$. 
	\item $x$: vector $(x_1,\dots,x_n)^\top$.
	\item $s_1$ : concentration of substrate 1 in $[g/l]$. In the example $s_1$ represents ammonium.
	\item $s_2$ : concentration of substrate 2 in $[g/l]$. In the example $s_2$ represents nitrite. 
	\item $s_3$ : concentration of substrate 3 in $[g/l]$. In the example $s_3$ represents nitrate.
	\item $s_{in}$: entry concentration of substrate $1$ in $[g/l]$. May depend on time $s_{in} = s_{in}(t)$.
	\item $s$:  vector $(s_1,s_2,s_3)$. Refered to as metabolites.
	\item $\I_i(t,x)$ : Interaction function of OTU $i \in \{1,\dots,n\}$.
	\item $\mu_i(s,x)$ : growth function of OTU $i \in \{1,\dots,n\}$.
	\item $\mu = (\mu_1(s,x), \dots, \mu_n(s,x))$ vector containing the growth function of every OTU.
	\item $D$: dilution rate of the continuous reactor in $[1/day]$. May depend on time $D=D(t)$.
	\item $y_i$: yield of grams of dry biomass formed per gram of substrate consumed of OTU $i$.
	\item $y$: vector $y_1,\dots,y_n$.
	\item $k_i$: yield of grams of substrate consumed per gram of biomass formed of OTU $i$ . The inverse of the biomass yield $k_i = \dfrac{1}{y_i}$.
	\item $k$: vector $k_1,\dots,k_n$.
	\item For integers $m_1$ and $m_2$ and $a \in \R$, $a_{m_1 \times m_2}$ represents a matrix of $m_1$ rows and $m_2$ columns with $a$ in every entry. 
	\item Let $m$ be an integer then $I_m$ is the identity matrix of size $m$.
	\item Let $S$ be a finite set with $m \in \N$ elements. Then $\vert S \vert := m$. 
	\item Given a vector $v=(v_1,\dots, v_n)\in \R^n$, the function $\diag(v)$ stands for:
	\begin{align}
	\begin{array}{rc}
	\diag:\R^n \rightarrow & \mathbb{M}_{n\times n}(\R)\\
	& \\
	v \rightarrow & \begin{pmatrix}
	v_1 & 0 & \dots & 0 \\
	0 & v_2 & \ddots & \vdots\\ 
	\vdots & \ddots & \ddots& 0 \\
	0 & \dots &0 & v_n 
	\end{pmatrix}
	\end{array} \label{diag_operator}
	\end{align}
\end{enumerate}

@@

The situation is described as simplified reactions \eqref{Reaction1} and \eqref{Reaction2}. The reactions are simplified in the sense that they do not attempt to represent a balanced chemical reaction, rather it represents the direction of the bioprocess and the proportions of different consumed and formed compounds of interest.

The terms $y_i$ are known as yields, they represent the quantity (in grams of dry biomass produced per gram of substrate) consumed for OTU $i$. In this work the unit grams of dry biomass per gram of substrate is used for yields. Furthermore, for each $ i \in [n]$, OTU $i$ is characterized by its process rate (also known as growth function) $\mu_i(s,x)$. Notice that for being as generic as possible, the growth rate may be a function of the whole state in order to model the influence of all OTU on the growth rates of others.
\begin{align}
\label{Reaction1} \tag{R G1} s_1 \stackrel{\mu_i(s,x)}{\longrightarrow} s_2 + y_ix_i \quad \forall i \in G_1 \\
\label{Reaction2} \tag{R G2} s_2\stackrel{\mu_i(s,x)}{\longrightarrow} s_3 + y_ix_i \quad \forall i \in G_2
\end{align}

@@

\begin{align} 
\label{system}
\begin{array}{cl}
\dot{x_i} =& \left(\mu_i(s,x) -D \right)x_i \quad \forall i \in G_1\\
\dot{x_i} =& \left(\mu_i(s,x) -D \right)x_i \quad \forall i \in G_2 \\
\dot{s_1} =& \displaystyle (s_{in}-s_1)D-\sum\limits_{i\in G_1} \frac{1}{y_i}\mu_i(s,x) x_i  \\
\dot{s_2} = & \displaystyle -s_2D+\sum\limits_{i\in G_1}\frac{1}{y_i}\mu_i(s,x)x_i 		-\sum\limits_{i\in G_2}\frac{1}{y_i}\mu_i(s,x) x_i  \\
\dot{s_3} =&  \displaystyle -s_3D+\sum\limits_{i\in G_2}\frac{1}{y_i}\mu_i(s,x) x_i 
\end{array}
\end{align}

@@

The proof can be seen in the supplementary material. A bound on the norm of $A$ is found such that every matrix $A$, respecting the bound, guaranties that $\Omega$ is a positively invariant set.

@@

Model \eqref{system} is restated in a more compact form using matrix notation.

\begin{align} A = \begin{bmatrix}
A_1\\A_2
\end{bmatrix}\end{align}
where $A_1$ is a matrix of dimension $n_1 \times n$ corresponding to the interaction coefficients of OTU $i \in G_1$. $A_2$ is a matrix of dimension $n_2 \times n$ rows corresponding to the interaction coefficients of OTU $i \in G_2$, and  $f(s) \in \R^n$ is such that 
\begin{align}
f_i(s) = \begin{cases}
\bar{\mu}_i \dfrac{s_1}{K_i + s_1} \quad \forall i \in G_1 \\
\bar{\mu}_i \dfrac{s_2}{K_i + s_2} \quad \forall i \in G_2
\end{cases}
\end{align}

@@

\begin{align}
\label{Eq1}	\dot{x} = & \diag(x)(\diag(f(s))(1_{n\times 1} + Ax) - D_{n\times 1}) \\
\label{Eq2}	\dot{s_1} = & (s_{in}-s_1)D + \begin{bmatrix}
-(1_{n_1\times 1} +A_1x) \\0
\end{bmatrix}^\top \diag(k) \diag(f(s))x \\
\label{Eq3}	\dot{s_2} = & -s_2D+
\begin{bmatrix}
1_{n_1\times 1} +A_1x \\ -(1_{n_2\times 1} +A_2x)
\end{bmatrix}^\top \diag(k) \diag(f(s))x \\	
\label{Eq4}	\dot{s_3} = & -s_3D+\begin{bmatrix}
0	\\(1_{n_2\times 1} +A_2x) 
\end{bmatrix}^\top \diag(k) \diag(f(s))x 
\end{align}

@@

\begin{defn}
	Given an equilibrium point $(x^{eq},s^{eq})$ of system \eqref{system}, then the set of non-active coordinates $\mathcal{J}\subset \{1,\dots,n\}$ is defined as:
	$\mathcal{J} = \{j_1, \dots, j_m\ : \, x^{eq}_{j_i} = 0 , \;i \in [m] \} $. $n_1^{act}$ and $n_2^{act}$ denote the number of positive entries of $x^{eq}$ of functional groups $G_1$ and $G_2$, respectively. $n^{act} = n-m$ denotes the total number of positive entries of $x^{eq}$. The active point $x^{act}\in \R^{n^{act}}$ is defined by the positive entries of $x^{eq}$. The active growths $f^{act}(s)$ and yields $k^{act}$ are defined analogously by the positive entries of $x^{eq}$. The active interactions $A^{act}$ is defined as the matrix $A$ without the $\mathcal{J}$ rows and columns. Let $M$ be the matrix defined by taking out the $\mathcal{J}$ columns of the identity matrix, then
	 \begin{align}
	 x^{eq} = Mx^{act} \\	
	 f^{act}(s) = M^{\top}f(s) \\
	 k^{act} = M^{\top}k \\
	 A^{act} = M^{\top}AM
	 \end{align}.

%	\begin{align}\label{EquilibriaFormula} (x^{eq})_i = \begin{cases}
%	x^{act}_i & \text{for } i < j_1\\
%	x^{act}_{i-k} & \text{for } j_{k}< i <j_{k+1}  \text{ and } k \in \{1,\dots,m-1\}\\
%	x^{act}_{i-m} & \text{for } j_{m}< i \\
%	0 & \text{for } i \in \{j_1, \dots, j_m\}
%	\end{cases} \end{align}

\end{defn}

@@

By replacing \eqref{EqSome} in equation \eqref{Eq2}:

\begin{align}
\label{s1Equilibria} (s_{in}-s_1)-	\begin{bmatrix}
1_{n_1^{act}\times 1} \\0_{n_2^{act}\times 1}
\end{bmatrix}^\top \diag (k^{act}) (A^{act})^{-1}(\diag(f^{act}(s))^{-1}D_{n^{act}\times 1}-1_{n^{act}\times 1}) = 0 
\end{align}

@@

\begin{align}
\label{s2Equilibria}-s_2+
	\begin{bmatrix}
1_{n_1^{act}\times 1} \\-1_{n_2^{act}\times 1}
\end{bmatrix}^\top\diag (k^{act})(A^{act})(\diag(f^{act}(s))^{-1}D_{n^{act}\times 1}-1_{n^{act}\times 1}) = 0
\end{align}

@@

In the case of ecological diagrams (ED) the pairs are regrouped such that the points in the set $\mathcal{P}$ have the same non-active coordinates. Ecological diagrams provide more information than operating diagrams, in the sense that one can infer the operating diagram from an ecological diagram. In other words, instead of representing areas where either CN, PN or WO take place, ranges of pairs $(s_{in},D)$ where species coexist are represented.

@@

Consider the case where $n_1 = 1$ and $n_2 = 1$. The operating diagrams where no interactions take place ($A = 0$) and a non-zero interaction matrix is presented. When $A=0$ algorithm \ref{StabilityAnalysis} is no longer valid (A is not invertible), nevertheless the stability analysis is much simpler and is given in the supplementary material section. The interaction matrices are shown in \ref{case1_no_interactions} and \ref{case1_interactions}, the rational behind the second choice was to force a . Parameters can be seen in table \ref{kinetic_parameters_case_study_1}, while

The operating diagram can be seen in figure \ref{OD_no_interactions}, note how partial nitrification (washout of $G_2$). When compared to \ref{OD_no_interactions} one notes that at low $s_{in}$ the curve dividing the PN zone from the CN zone are alike. But since more $s_{in}$ means more $x_1$, at high $s_{in}$ the affine part $(1+ a_{21}x_1 +a_{22}x_2)$ of the growth function of $x_2$ plays a bigger role than substrate limitation ($\frac{s_2}{s_2+K_2}$).

@@

\begin{figure}[h]
	\centering
	\begin{subfigure}[b]{0.45\textwidth}
		\includegraphics[width = \textwidth]{Stability/Interactions_case_study_1_no_interactions}
		\caption{Interaction matrix of model \eqref{system} with no interactions.}
		\label{case1_no_interactions}
	\end{subfigure}
	~
	\begin{subfigure}[b]{0.45\textwidth}
		\includegraphics[width = \textwidth]{Stability/Interactions_case_study_1}
		\caption{Interaction matrix of model \eqref{system}.}
		\label{case1_interactions}
	\end{subfigure}
	\caption{Interaction matrices. Note how the presence of $x_1$ affects very negatively $x_2$ in \eqref{case1_interactions}. The terms in the diagonal entries of the matrix represents intraspecies interactions, while the terms off thepo
		 diagonal represent the interspecies interactions.}
\end{figure}

\begin{figure}[h]
	\centering
	\begin{subfigure}[b]{0.45\textwidth}
		\includegraphics[width = \textwidth]{Stability/OD_case_study_1_no_interactions}
		\caption{Operating diagram of model \eqref{system} with no interactions (interaction matrix represented by \eqref{case1_no_interactions}).}
		\label{OD_no_interactions}
	\end{subfigure}
~
	\begin{subfigure}[b]{0.45\textwidth}
	\includegraphics[width = \textwidth]{Stability/OD_case_study_1}
	\caption{Operating diagram of model \eqref{system} with interactions represented by \eqref{case1_interactions}.}
	\label{OD_interactions}
	\end{subfigure}
	\caption{Note how \ref{OD_interactions} has a much larger zone where partial nitrification takes place. This is due to the negative interaction of $x_1$ on $x_2$.}
\end{figure} 
\clearpage

@@

 The case study 2 is based on Dumont \textit{et al.}\cite{Dumont2016} model parameters. They proposed a distribution of parameters obtained from a Bayesian estimation method. Their fit describes well the dynamics of the two most abundant OTU in each functional group, but it still fails to capture the measured substrates dynamics. Kinetic parameters of case study 2 can be seen in table \ref{kinetic_parameters}. The estimated interaction matrix is shown in figure \ref{Interaction_1}. A second matrix is presented, which is obtained by the sign change of coefficient $a_{11}$ (figure \ref{Interaction_2}), and finally a third one is obtained by using a positive value for $a_{12}$ (figure \ref{Interaction_3}).

@@

\begin{figure}[ht]
	\centering
	\begin{subfigure}[b]{0.32\textwidth}
	\includegraphics[width=\textwidth]{Stability/ED_parameters_Dumont}
	\caption{Original parameters ecological diagram.}
	\label{ED 1}
	\end{subfigure}
~
	\begin{subfigure}[b]{0.32\textwidth}
	\includegraphics[width=\textwidth]{Stability/ED_parameters_modified}
	\caption{Modified interaction matrix ecological diagram.}
	\label{ED 2}
	\end{subfigure}
~
\begin{subfigure}[b]{0.32\textwidth}
	\includegraphics[width=\textwidth]{Stability/ED_parameters_modified_2}
	\caption{Modified interaction matrix ecological diagram.}
	\label{ED 3}
\end{subfigure}

@@

The ecological diagrams are presented in figure \ref{ecological_diagrams}. Comparing figure \ref{ED 1} with \ref{ED 2} one can see a very different situation. Note how every zone in figure \ref{ED 2} has two stable equilibria, meaning that the outcome of the system is determined by its initial conditions, particularly interesting is the green zone where either $x_1,x_3$ coexist or $x_2$ alone remains, because in operational terms this means that either PN or CN may take place.

One can see that coexistence in the same functional group is never attained in figure \ref{ED 1}, whereas in figure \ref{ED 2} $x_1$ and $x_2$, both AOB, coexist in either partial or complete nitrification. That means that the model is able to overcome the competitive exclusion principle \cite{lobry2017chemostat} (CEP) paradigm. The CEP roughly states that if two species are growing on the same limiting resource, and their growth laws only depend non decreasingly on the resource, then only one of them will survive in the long run. This is interesting in light of reports on wastewater treatment plants where coexistence between species in nitrifying reactors has been shown \cite{Wagner2002}, thus implying that a more complex growth law (as shown in here) or model structure involving other biological processes is required to include microbial diversity in mathematical models.

\subsection{Remarks}

The model here presented serves to illustrate that by considering a more complex growth rate that tries to model ecological interactions one might explain differences in reactors operating under similar conditions. It also shows a new mechanism by which the CEP no longer holds and where multiple stable equilibria may appear. Since the gLV model discussed (and calibrated in \cite{Dumont2016}) fails to completely capture the dynamics observed in the chemostat experiments, the next section proposes a new model to study interactions.

@@

Given a vector $(v_1,\dots,v_n)^\top$ the interaction function $\I$ is noted as:
\begin{align}
\begin{array}{rc}
\I: \R_+^n \rightarrow & \R_+^n\\
& \\
v \rightarrow & \begin{pmatrix}
I_1(v) \\
I_2(v) \\ 
\vdots  \\
I_n(v)
\end{pmatrix}
\end{array}
\end{align}

@@

\begin{hypo}
	The interaction function $\I$ previously defined satisfies:
	\begin{enumerate}
		\item $	\I \left( (0,\dots,0)^\top \right) = (1,\dots,1)^\top $
		\item Let $\Omega$ be an open subset of $ \R^n $ then $\I \in C^1(\Omega)$.
	\end{enumerate} 
\end{hypo}

@@

The feedback control $v(t)$, will be of the form $v(t) = -R^{-1}\tilde{B}^\top(t)\left(\tilde{P}(t)x(t)-s_f(t)\right)$. Where matrix $P(t)$ and vector $s_f(t)$ solve differential equations. In the appendix it is proved that because of the structure of the system one only needs to calculate $2n$ differential equations for the synthesis of the control, instead of $(n+3)^2 + (n+3)$ that would imply the direct application of the method. Which renders the method- at least theoretically- scalable for a growing number of OTU.

>>>>

Notations used throughout the article:
\begin{enumerate}
	\item $n$: the number of OTU considered.
	\item $n_i, \, i \in \{1,2\}$: the number of OTU in functional group $G_i$. In the example $G_1$ corresponds to Ammonium oxidizing bacteria (AOB) and $G_2$ corresponds to Ammonium oxidizing bacteria (NOB).
	\item Let $m$ be an interger then $[m]:=\{1,\dots,n \}$. 
	\item $x_i$ :  is the concentration of OTU $i$ measured in $[g/l]$. $i \in [n]$. 
	\item $x$: vector $(x_1,\dots,x_n)^\top$.
	\item $s_1$ : concentration of substrate 1 in $[g/l]$. In the example $s_1$ represents ammonium.
	\item $s_2$ : concentration of substrate 2 in $[g/l]$. In the example $s_2$ represents nitrite. 
	\item $s_3$ : concentration of substrate 3 in $[g/l]$. In the example $s_3$ represents nitrate.
	\item $s_{in}$: entry concentration of substrate $1$ in $[g/l]$. May depend on time $s_{in} = s_{in}(t)$.
	\item $s$:  vector $(s_1,s_2,s_3)$. Refered to as metabolites.
	\item $\I_i(t,x)$ : Interaction function of OTU $i \in \{1,\dots,n\}$.
	\item $\mu_i(s,x)$ : growth function of OTU $i \in \{1,\dots,n\}$.
	\item $\mu = (\mu_1(s,x), \dots, \mu_n(s,x))$ vector containing the growth function of every OTU.
	\item $D$: dilution rate of the continuous reactor in $[1/day]$. May depend on time $D=D(t)$.
	\item $y_{i}$: yield of grams of OTU $i$ formed per gram of substrate consumed. 
	\item $y_{s_i/x_j}$: yield of grams of substrate $s_i$ consumed/produced per gram of OTU $j$ formed. If negative it represents consumption, if positive it represents production.
	\item $Y$: matrix containing all yields such that $Y_{ij} = y_{s_i/x_j}$.
	\item For integers $m_1$ and $m_2$ and $a \in \R$, $a_{m_1 \times m_2}$ represents a matrix of $m_1$ rows and $m_2$ columns with $a$ in every entry. 
	\item Let $m$ be an integer then $I_m$ is the identity matrix of size $m$.
	\item Let $M$ be a matrix, then $M_{i\bullet}$ represents the $i$-th row of matrix $M$.
	\item Let $S$ be a finite set with $m \in \N$ elements. Then $\vert S \vert := m$. 
	\item Given a vector $v=(v_1,\dots, v_n)\in \R^n$, the function $\diag(v)$ stands for:
	\begin{align}
	\begin{array}{rc}
	\diag:\R^n \rightarrow & \mathbb{M}_{n\times n}(\R)\\
	& \\
	v \rightarrow & \begin{pmatrix}
	v_1 & 0 & \dots & 0 \\
	0 & v_2 & \ddots & \vdots\\ 
	\vdots & \ddots & \ddots& 0 \\
	0 & \dots &0 & v_n 
	\end{pmatrix}
	\end{array} \label{diag_operator}
	\end{align}
\end{enumerate}

@@

The situation is described as simplified reactions \eqref{Reaction1} and \eqref{Reaction2}. The reactions are simplified in the sense that they do not attempt to represent a balanced chemical reaction, rather they represent the direction of the bioprocess and the proportions of different consumed and formed compounds of interest. The terms $y_i$ are known as yields, they represent the quantity of g of biomass produced per g of substrate consumed by  OTU $i$. For example in the case of reaction \eqref{Reaction1}, one gram of $s_1$ is consumed, one gram of $s_2$ and $y_{x_i/s_1}$ grams of dry biomass of OTU $i$ are produced. 
\begin{align}
\label{Reaction1} \tag{R G1} s_1 \stackrel{\mu_i(s,x)}{\longrightarrow} s_2 + y_ix_i \quad \forall i \in G_1 \\
\label{Reaction2} \tag{R G2} s_2\stackrel{\mu_i(s,x)}{\longrightarrow} s_3 + y_ix_i \quad \forall i \in G_2
\end{align}
However for expressing the system of differential equations further below, the terms $y_{s_i/x_j}$ are used. They express the grams of substrate $s_i$ consumed (negative sign) or produced (positive sign) per gram of OTU $j$ formed. They are related to $y_i$ as seen in table \ref{yields table}. This defines the stoichiometry matrix $Y \in \R^{3 \times n}$, such that $Y_{ij} = y_{s_i/x_j}$.
\begin{table}[h]
	\centering
	\begin{tabular}{|l|l|l|}
		\hline
		\multicolumn{1}{|c|}{\begin{tabular}[c]{@{}c@{}}Yields per \\ Biomass formed\end{tabular}} & \multicolumn{1}{c|}{$j \in G_1$} & \multicolumn{1}{c|}{$j \in G_2$} \\ \hline   \rule{0pt}{3.5ex}
		$y_{s_1/x_j}$ & $-\dfrac{1}{y_j}$ & $0$ \\  \hline   \rule{0pt}{3.5ex}
		$y_{s_2/x_j}$ & $\dfrac{1}{y_j}$  & $-\dfrac{1}{y_j}$ \\ \hline   \rule{0pt}{3.5ex}
		$y_{s_3/x_j}$ & $0$  & $\dfrac{1}{y_j}$  \\ \hline
	\end{tabular}
	\caption{Relationship of $y_{s_i/x_j}$ with $y_j$.}
		\label{yields table}
\end{table}

Furthermore, for each $ i \in [n]$, OTU $i$ is characterized by its process rate (also known as growth function) $\mu_i(s,x)$. Notice that for being as generic as possible, the growth rate may be a function of the whole state in order to model the influence of all OTU on the growth rates of others.

@@

\begin{align} 
\label{system}
\begin{array}{cl}
\dot{x_i} =& \left(\mu_i(s,x) -D \right)x_i \quad \forall i \in [n] \\
\dot{s_1} =& \displaystyle (s_{in}-s_1)D+\sum\limits_{i\in G_1} y_{s_1/x_i}\mu_i(s,x) x_i  \\
\dot{s_2} = & \displaystyle -s_2D+\sum\limits_{i\in [n]}y_{s_2/x_i}\mu_i(s,x)x_i  \\
\dot{s_3} =&  \displaystyle -s_3D+\sum\limits_{i\in G_2}y_{s_3/x_i}\mu_i(s,x) x_i 
\end{array}
\end{align}

@@

The proof can be seen in the supplementary material. A bound on the norm of $A$ is found such that every matrix $A$ respecting the bound, guaranties that $\Omega$ is a positively invariant set.

@@

Model \eqref{system} is restated in a more compact form using matrix notation. Let $f(s)$ be such that,
%
%\begin{align} A = \begin{bmatrix}
%A_1\\A_2
%\end{bmatrix}\end{align}
%where $A_1$ is a matrix of dimension $n_1 \times n$ corresponding to the interaction coefficients of OTU $i \in G_1$. $A_2$ is a matrix of dimension $n_2 \times n$ rows corresponding to the interaction coefficients of OTU $i \in G_2$, and  $f(s) \in \R^n$ is such that 
\begin{align}
f_i(s) = \begin{cases}
\bar{\mu}_i \dfrac{s_1}{K_i + s_1} \quad \forall i \in G_1 \\
\bar{\mu}_i \dfrac{s_2}{K_i + s_2} \quad \forall i \in G_2
\end{cases}
\end{align}

Then $\mu(x,s) = \diag(f(s))(1_{n\times 1} + Ax)$

@@

%\begin{align}
%\label{Eq1}	\dot{x} = & \diag(x)(\diag(f(s))(1_{n\times 1} + Ax) - D_{n\times 1}) \\
%\label{Eq2}	\dot{s_1} = & (s_{in}-s_1)D + \begin{bmatrix}
%-(1_{n_1\times 1} +A_1x) \\0
%\end{bmatrix}^\top \diag(k) \diag(f(s))x \\
%\label{Eq3}	\dot{s_2} = & -s_2D+
%\begin{bmatrix}
%1_{n_1\times 1} +A_1x \\ -(1_{n_2\times 1} +A_2x)
%\end{bmatrix}^\top \diag(k) \diag(f(s))x \\	
%\label{Eq4}	\dot{s_3} = & -s_3D+\begin{bmatrix}
%0	\\(1_{n_2\times 1} +A_2x) 
%\end{bmatrix}^\top \diag(k) \diag(f(s))x 
%\end{align} 
\begin{align}
\label{Eq1}	\dot{x} = & \diag(\mu(x,s) - D_{n\times 1})x \\
\label{Eq2}	\dot{s_1} = & (s_{in}-s_1)D + Y_{1\bullet} \diag(\mu(x,s))x \\
\label{Eq3}	\dot{s_2} = & -s_2D+
Y_{2\bullet} \diag(\mu(x,s))x\\	
\label{Eq4}	\dot{s_3} = & -s_3D+Y_{3\bullet} \diag(\mu(x,s))x 
\end{align}

@@

\begin{defn}
	Given an equilibrium point $(x^{eq},s^{eq})$ of system \eqref{system}, then the set of non-active coordinates $\mathcal{J}\subset \{1,\dots,n\}$ is defined as:
	$\mathcal{J} = \{j_1, \dots, j_m\ : \, x^{eq}_{j_i} = 0 , \;i \in [m] \} $. $n_1^{act}$ and $n_2^{act}$ denote the number of positive entries of $x^{eq}$ of functional groups $G_1$ and $G_2$, respectively. $n^{act} = n-m$ denotes the total number of positive entries of $x^{eq}$. The active point $x^{act}\in \R^{n^{act}}$ is defined by the positive entries of $x^{eq}$. Analogously, the functions $f^{act}(s)$ and $\mu^{act}(x,s)$ are defined by the positive entries of $x^{eq}$. The active interactions $A^{act}$ is defined as the matrix $A$ without the $\mathcal{J}$ rows and columns. The active stoichiometry matrix $Y^{act}$ is the matrix $Y$ without the $\mathcal{J}$ columns. In summary, let $M$ be the matrix defined by taking out the $\mathcal{J}$ columns of the identity matrix of size $n$, then
	 \begin{align}
	 x^{eq} = Mx^{act} \\	
	 f^{act}(s) = M^{\top}f(s) \\
	 \mu^{act}(x,s) = M^{\top}\mu(x,s)\\
	 Y^{act} = YM \\
	 A^{act} = M^{\top}AM
	 \end{align}.   
\end{defn}

@@

By replacing $x^{eq} = Mx^{act}$ in equation \eqref{Eq2}:

\begin{align}
 (s_{in}-s_1)D + Y_{1\bullet} \diag(\mu(x,s))Mx^{act}  = 0 \\
(s_{in}-s_1)D + Y^{act}_{1\bullet} \diag(\mu^{act}(x,s))x^{act}  = 0 \\ 
(s_{in}-s_1)D + Y^{act}_{1\bullet} \diag(D_{n^{act}\times 1})x^{act}  = 0  \\
(s_{in}-s_1) + Y^{act}_{1\bullet} x^{act}  = 0  \\
\label{s1Equilibria} (s_{in}-s_1) + Y^{act}_{1\bullet}(A^{act})^{-1}(\diag(f^{act}(s))^{-1}D_{n^{act}\times 1} - 1_{n^{act}\times 1}) = 0
\end{align}

@@

\begin{align}
\label{s2Equilibria}-s_2+ Y^{act}_{2\bullet}(A^{act})^{-1}(\diag(f^{act}(s))^{-1}D_{n^{act}\times 1} - 1_{n^{act}\times 1}) = 0
\end{align}

@@

In the case of ecological diagrams (ED) the pairs are regrouped such that the points in the set $\mathcal{P}$ have the same non-active coordinates. In other words, instead of representing areas where either CN, PN or WO take place, ranges of pairs $(s_{in},D)$ where species coexist are represented.	ED provide more information than OD, in the sense that one can deduce the latter from the former.

@@

Consider the case where $n_1 = 1$ and $n_2 = 1$. The operating diagrams when no interactions take place ($A = 0$) and a non-zero interaction matrix are presented. When $A=0$ algorithm \ref{StabilityAnalysis} is no longer valid (A is not invertible), nevertheless the stability analysis is much simpler and is given in the supplementary material section. The interaction matrices are shown in Figures \ref{case1_no_interactions} and \ref{case1_interactions}, the rational behind the second choice was to force a very strong interaction of $x_1$ on $x_2$ and observe its effects. The rest of parameters can be seen in Table \ref{kinetic_parameters_case_study_1}.

The operating diagrams can be seen in Figure \ref{OD_case_1}, note how partial nitrification (washout of $G_2$) of Figure \ref{OD_interactions} is much bigger when compared to Figure \ref{OD_no_interactions}. The shape of the PN region in \ref{OD_interactions} is somewhat unintuitive, because at a constant dilution rate (0.24 $day^{-1}$ for example) and an increasing $s_{in}$, one passes from a PN zone, to a CN zone, and then back again to a PN zone. The explanation lies in the fact that $x_1$ also increases with $s_{in}$, and the affine part $(1+ a_{21}x_1 +a_{22}x_2)$ of the growth function of $x_2$ plays a bigger role than substrate limitation ($\frac{s_2}{s_2+K_2}$).

@@

\begin{figure}[h]
	\centering
	\begin{subfigure}[b]{0.45\textwidth}
		\includegraphics[width = \textwidth]{Stability/Interactions_case_study_1_no_interactions}
		\caption{Interaction matrix of model \eqref{system} with no interactions.}
		\label{case1_no_interactions}
	\end{subfigure}
	~
	\begin{subfigure}[b]{0.45\textwidth}
		\includegraphics[width = \textwidth]{Stability/Interactions_case_study_1}
		\caption{A non-zero interaction matrix of model \eqref{system}.}
		\label{case1_interactions}
	\end{subfigure}
	\caption{Interaction matrices. Note how the presence of $x_1$ affects very negatively $x_2$ in Figure \ref{case1_interactions}, with respect to other interactions. The terms in the diagonal entries of the matrix represents intraspecies interactions, while the terms off the diagonal represent the interspecies interactions.}
\end{figure}

\begin{figure}[h]
	\centering
	\begin{subfigure}[b]{0.45\textwidth}
		\includegraphics[width = \textwidth]{Stability/OD_case_study_1_no_interactions}
		\caption{Operating diagram of model \eqref{system} with no interactions (interaction matrix represented by figure \ref{case1_no_interactions}).}
		\label{OD_no_interactions}
	\end{subfigure}
~
	\begin{subfigure}[b]{0.45\textwidth}
	\includegraphics[width = \textwidth]{Stability/OD_case_study_1}
	\caption{Operating diagram of model \eqref{system} with interactions represented by figure \ref{case1_interactions}.}
	\label{OD_interactions}
	\end{subfigure}
	\caption{Note how \ref{OD_interactions} has a much larger zone where partial nitrification takes place. This is due to the negative interaction of $x_1$ on $x_2$.}
	\label{OD_case_1}
\end{figure} 
\clearpage

@@

 The case study 2 is based on Dumont \textit{et al.}\cite{Dumont2016} model parameters. They proposed a distribution of parameters obtained from a Bayesian estimation method. Their fit describes well the dynamics of the two most abundant OTU in each functional group, but it still fails to capture the measured substrates dynamics. Kinetic parameters of case study 2 can be seen in Table \ref{kinetic_parameters}. The estimated interaction matrix is shown in Figure \ref{Interaction_1}. A second matrix is presented, which is obtained by the sign change of coefficient $a_{11}$ (Figure \ref{Interaction_2}), and finally a third one is obtained by using a positive value for $a_{12}$ (Figure \ref{Interaction_3}). The idea is to show that qualitatively different outcomes can be obtained by changing one interaction at a time.

@@

\begin{figure}[h]
	\centering
	\begin{subfigure}[t]{0.32\textwidth}
	\includegraphics[width=\textwidth]{Stability/ED_parameters_Dumont}
	\caption{Original parameters ecological diagram.}
	\label{ED 1}
	\end{subfigure}
~
	\begin{subfigure}[t]{0.32\textwidth}
	\includegraphics[width=\textwidth]{Stability/ED_parameters_modified}
	\caption{Modified interaction matrix ecological diagram. In the legend 1) and 2) represent the two different stable equilibria in each zone.}
	\label{ED 2}
	\end{subfigure}
~
\begin{subfigure}[t]{0.32\textwidth}
	\includegraphics[width=\textwidth]{Stability/ED_parameters_modified_2}
	\caption{Modified interaction matrix ecological diagram.}
	\label{ED 3}
\end{subfigure}

@@

The ecological diagrams are presented in Figure \ref{ecological_diagrams}, where the legend indicates the species surviving in the zone of the respective colour. In Figure \ref{ED 2}, the system exhibits bi-stability (it is represented in numbering as 1) and 2) of the different possible equilibriums) . Comparing Figures \ref{ED 1} with \ref{ED 2} one can see a very different situation. Note how every zone in figure \ref{ED 2} has two stable equilibria, meaning that the outcome of the system is determined by its initial conditions, particularly interesting is the green zone where either $x_1,x_3$ coexist or $x_2$ only remains, because in operational terms this means that either PN or CN may take place.

One can see that coexistence in the same functional group is never attained in Figure \ref{ED 1} and \ref{ED 2}, whereas in figure \ref{ED 3} $x_1$ and $x_2$, both AOB, coexist in either partial or complete nitrification. That means that the competitive exclusion principle \cite{lobry2017chemostat} (CEP) does not hold. The CEP roughly states that if two species are growing on the same limiting resource, and their growth laws only depend non decreasingly on the resource, then only one of them will survive in the long run. This is interesting in light of reports on wastewater treatment plants where coexistence between species in nitrifying reactors has been shown \cite{Wagner2002}, thus implying that a more complex growth law (as shown in here) or model structure involving other biological processes is required to include microbial diversity in mathematical models.

\section*{Remarks}

Model \eqref{system} serves to illustrate that by considering a more complex growth rate that tries to model ecological interactions one might explain differences in reactors operating under similar conditions. It also shows a new mechanism by which the CEP no longer holds and which explains how multiple stable equilibria may appear. Since the gLV model discussed fails to completely capture the dynamics observed in the chemostat experiments \cite{Dumont2016}, the next section proposes a new approach to study interactions.

@@

Given a vector $(v_1,\dots,v_n)^\top$ the interaction function $\I$ is denoted as:
\begin{align}
\begin{array}{rc}
\I: \R_+^n \rightarrow & \R_+^n\\
& \\
v \rightarrow & \begin{pmatrix}
I_1(v) \\
I_2(v) \\ 
\vdots  \\
I_n(v)
\end{pmatrix}
\end{array}
\end{align}

@@

\begin{hypo}
	The interaction function $\I$ previously defined satisfies:
	\begin{enumerate}
		\item $	\I \left( (0,\dots,0)^\top \right) = (1,\dots,1)^\top $
		\item There is an open set $\Omega \subset \R^n $ such that $\I \in C^1(\Omega)$.
	\end{enumerate} 
\end{hypo}

@@

The feedback control $v(t)$, will be of the form $v(t) = -R^{-1}\tilde{B}^\top(t)\left(\tilde{P}(t)x(t)-s_f(t)\right)$. Where matrix $P(t)$ and vector $s_f(t)$ solve differential equations. In the appendix it is proved that because of the structure of the system one only needs to calculate $2n$ differential equations for the synthesis of the control, instead of $(n+3)^2 + (n+3)$ that would imply the direct application of the method, which renders the method- at least theoretically- scalable for a growing number of OTU.
